Structuring College Access

### 2025-10-09 aefp/aefp_geomarket.qmd
added cerpp proposal

<!--  
we argue that utilizing Geomarket filters to purchase student lists can contribute to socioeconomic and racial inequality on college opportunity.

Using data about past correlations -- here, Geomarket borders created with reference to 1980 SAT score sending behaviors -- to make future recommendations amplifies the effects of historical structural inequalities. 
-->

>>>>

```{=html}
<!--  
we argue that utilizing Geomarket filters to purchase student lists can contribute to socioeconomic and racial inequality on college opportunity.

Using data about past correlations -- here, Geomarket borders created with reference to 1980 SAT score sending behaviors -- to make future recommendations amplifies the effects of historical structural inequalities. 
-->
```

### 2025-10-09 cerpp/cerpp_geomarket.qmd
added cerpp proposal

<!--  
\thispagestyle{empty} 
\newpage 
\setcounter{page}{1}
Please submit a detailed 2-page précis that includes a brief discussion of the topic, pertinent literature and
theory, research methods, findings, and study significance

@@

HOW TO CONNECT WITH THE THEME
- sociology of organizations knows that regulating an industry requires understanding the organizational field, which includes third-party vendors and products. these third party vendors and agents structure college access in ways that benefit historically privileged groups and contribute to accumulation of capital
- we must understand the field before we can regulate it; CFPB has 
Education policy has been dominated by economists. economics doesn't have the theoretical concepts necessary 
-->

@@

@@

**BACKGROUND.** In 1983, Zemsky and Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board*, @RN4982 developed the Market Segment Model and created "Geomarkets." The Market Segment Model predicts how student demand for a particular college varies across local geographic markets based on the characteristics of students in those local markets.

Geomarkets are geographic areas that divide states and large metropolitan areas into "community-based enrollment markets" [@RN4982, p. 14] for recruiting purposes. @fig-nyc-geomarkets shows New York City area Geomarkets. The Market Segment Model and Geomarkets are the basis for the College Board Enrollment Planning Service (EPS), founded in 1984 and still active today. EPS software recommends which Geomarkets a college should recruit from and which high schools within targeted Geomarkets a college should visit.[^1]

[^1]: In 1995, 37% of 4-year public institutions and 49% of 4-year private institutions used EPS [@RN5021].

Geomarkets were incorporated into the *College Board* "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education for over 50 years [@belkin2019-studata; @list_biz].[^2] Lists contain contact information for prospective students. The Student Search Service database consists of College Board test-takers. Colleges pay a fee for each prospect (e.g., \$0.50). Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to target prospects living in particular Geomarkets.^[An analysis of 830 student lists purchased by 14 public universities found that 16% of purchases filtered on Geomarket [@RN5012].]

[^2]: @RN4895 reported that 87% of private and 86% of public four-year institutions purchase lists.

**THEORY & RQs.** We analyze the Market Segment Model and Geomarkets using scholarship on quantification [e.g., @RN5003], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. The 1980 SAT score-sending data analyzed by @RN4982 can be conceived as "training data," which was used to define four "market segments" of students: "local" students, who send most SAT scores to colleges in their Geomarket; "in-state" students, who primarily send scores to colleges in their state; "regional" students, who send most scores to colleges outside their state but in their region (e.g., New England); and "national" students. @RN4982 analyzed the student characteristics correlated with student market segment and concluded that student demand was primarily a function of class. The overarching thesis of the Market Segment Model is homophily, the idea that similar colleges compete with one another for similar students. @RN4982[p. 42-45] states that selective colleges primarily draw from students in the "regional" and "national" segments and, therefore, their recruiting visits and student list purchases should focus on Geomarkets with large numbers of affluent, highly educated households.

When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality [@RN4794]. A snapshot of student demand in 1980, itself a consequence of historical structural inequality, became inscribed into the EPS and Student Search Service products used by colleges to plan and implement recruiting strategy. The result is a supply-side that reinforces structural inequalities observed on the demand-side. We address two research questions:

@@

**SIGNIFICANCE.** Scholarship on quantification in the sociology of education [e.g.,  @RN2356;@RN5005] has not analyzed third-party products that colleges use to classify and identify prospective students. More broadly, the concept of "organizational fields" includes all organizations that "constitute a recognized area of institutional life"  [@RN527, p. 143]. In the enrollment management industry, third-party vendors and their products structure college access in ways that funnel certain kinds of students to certain kinds of institutions. Higher education policy over the last 20 years -- dominated by economists -- has focused on regulating consumers (students) and direct providers (colleges). However, federal regulatory agencies (e.g., FTC, CFPB) are increasingly interested in regulating the entire organizational field because they know that third-party oligopolies can wield market power over consumers and providers [@philippon2019great]. Therefore, scholarship from sociology should guide which problems and actors become the focus of future federal regulation.

**DATA SOURCES**. To answer RQ1, we utilize census tract-level data from the 1980 Decennial Census, 2000 Decennial Census, and 2020 5-year American Community Survey (ACS). Variables of interest include: race/ethnicity; mean and median household income; percent of households below poverty line; and BA attainment of people aged 25+. We obtained shapefiles for College Board Geomarkets from a 2012 *R-bloggers* post. We assigned census tracts to Geomarkets by implementing a partial spatial join.

To answer RQ2, we issued public records requests about student lists purchased by 14 public universities from 2016 through 2020. For 414 College Board student list purchases, associated with about 2.6 million prospects, we received both (1) the order summary -- showing which search filters were utilized -- and (2) the de-identified prospect-level data, including race, ethnicity, zip code, and high school. @fig-asu-order-448922 shows the partial order summary, which we converted to tabular data, for a purchase by Arizona State University that filtered on the 2020 high school graduating class, PSAT scores of 1070 to 1180, and selected Geomarkets.

@@

**PRELIMINARY FINDINGS**. For RQ1, Table \ref{tab:chi_eps_table_summary} shows racial and socioeconomic characteristics for Geomarkets around Chicago. @fig-socal-hispanic is a screenshot of an interactive choropleth map of Los Angeles, showing Geomarket borders and the tract-level %Hispanic in 2020. @fig-socal-med-inc shows tract-level median household income around Los Angeles.

Results for RQ2 show how many purchased prospects from each racial/ethnic group live in each Geomarket. These prospects would have been excluded had the purchase filtered on Geomarkets and not selected that Geomarket. @tbl-philly-sim-count, @tbl-philly-sim-col-pct, and @tbl-philly-sim-row-pct, respectively, show counts, column percentages, and row percentages for the Philadelphia metropolitan area, based on a purchase that filtered on students from the 2020 high school class who scored between 1070 and 1180 on the PSAT. @tbl-philly-sim-col-pct shows that "PA5--Philadelphia County" contains 15% of all purchased prospect profiles but 44% of Black prospects, indicating that opportunity for Black students is sensitive to the inclusion of this Geomarket. @fig-philly-black-prospects is a screenshot of a (very) preliminary interactive map for this purchase, with marker size indicating the number of purchased prospect profiles who attended that high school and separate markers by race/ethnicity.

@@

\newpage
\setcounter{page}{1}
\renewcommand{\thepage}{A-\arabic{page}}

# Appendix

```{r socal_eps_table_summary, echo=FALSE, results='asis', eval = FALSE}
# Generate the Los Angeles (socal) table with Bookdown label
socal_table <- format_eps_table(
  df = socal_eps_table, 
  years = c(2020), 
  metro_area = "Los Angeles", 
  format = "latex",
  label = "socal_eps_table_summary"  # This label matches the reference in your text
)

# Output the table directly
cat(socal_table)
```

```{r philly_eps_table_summary, echo=FALSE, results='asis', eval = FALSE}
# Wrap the table in the landscape environment
cat("\\begin{landscape}\n")

# Create the table for Philadelphia
philly_table <- format_eps_table(
  df = philly_eps_table, 
  years = c(2000, 2020), 
  metro_area = "Philadelphia metropolitan area", 
  format = "latex",
  label = "philly_eps_table_summary"  # This label matches the reference in your text  
)

# Output the tables directly
cat(philly_table)
cat("\n\\end{landscape}")
```

```{r chi_eps_table_summary, echo=FALSE, results='asis'}
# Wrap the table in the landscape environment
cat("\\begin{landscape}\n")

# Create the table for Philadelphia
chi_table <- format_eps_table(
  df = chi_eps_table, 
  years = c(2000, 2020), 
  metro_area = "Chicago-land", 
  format = "latex",
  label = "chi_eps_table_summary"  # This label matches the reference in your text  
)

# Output the tables directly
cat(chi_table)
cat("\n\\end{landscape}")
```

```{r create_eps_sim_tables, echo = FALSE, results='asis', message = FALSE, warning = FALSE}
# Load required packages and input data
load(file = file.path('.', '..', 'results', 'tables', 'philly_asu_eps_lists.RData'))
# str(philly_order448440_psat1270_1520)
# str(philly_order448922_psat1070_1180)

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)  # For adding commas to large numbers

# Function to format three separate tables
format_three_tables <- function(input_list) {
  # Extract individual data frames from the input list
  count_table <- input_list$count_table %>% 
    rename(all = stu_all, race_known = stu_race_known, white = stu_white, asian = stu_asian, black = stu_black, hispanic = stu_hispanic, two_races = stu_tworaces) %>%
    mutate(across(where(is.numeric), ~ if_else(. > 1000, comma(.), as.character(.))))  # Add commas to large numbers

  # Modify race_known to be a column percentage of the race_known total
  col_pct_table <- input_list$col_pct_table %>%
    rename(all = stu_all, race_known = stu_race_known, white = col_pct_white, asian = col_pct_asian, black = col_pct_black, hispanic = col_pct_hispanic, two_races = col_pct_tworaces) %>%
    select(-all) %>%
    mutate(
      race_known = (race_known / sum(race_known)) * 100,  # Convert race_known to a column percentage of the total
      across(-eps_codename, ~ round(., 1))  # Round percentage values to one decimal point
    )

  # Set race_known to always be 100% in row_pct_table
  row_pct_table <- input_list$row_pct_table %>%  
    rename(all = stu_all, race_known = stu_race_known, white = row_pct_white, asian = row_pct_asian, black = row_pct_black, hispanic = row_pct_hispanic, two_races = row_pct_tworaces) %>%
    select(-all) %>%
    mutate(
      race_known = 100.0,  # Set race_known to 100%
      across(-eps_codename, ~ round(., 1))  # Round percentage values to one decimal point
    )

  # Format and display count table with caption
  count_table_output <- count_table %>%
    kbl(format = "latex", booktabs = TRUE, align = c("l", rep("r", ncol(count_table) - 1))) %>%
    kable_styling(latex_options = c("striped", "hold_position"), position = "left", font_size = 10) %>%
    row_spec(0, bold = TRUE, background = "#D3D3D3")

  # Format and display column percentage table with caption
  col_pct_table_output <- col_pct_table %>%
    kbl(format = "latex", booktabs = TRUE, align = c("l", rep("r", ncol(col_pct_table) - 1))) %>%
    kable_styling(latex_options = c("striped", "hold_position"), position = "left", font_size = 10) %>%
    row_spec(0, bold = TRUE, background = "#D3D3D3")

  # Format and display row percentage table with caption
  row_pct_table_output <- row_pct_table %>%
    kbl(format = "latex", booktabs = TRUE, align = c("l", rep("r", ncol(row_pct_table) - 1))) %>%
    kable_styling(latex_options = c("striped", "hold_position"), position = "left", font_size = 10) %>%
    row_spec(0, bold = TRUE, background = "#D3D3D3")

  # Return formatted tables
  list(count_table_output, col_pct_table_output, row_pct_table_output)
}

# Example usage
formatted_tables <- format_three_tables(philly_order448922_psat1070_1180)

# Print formatted tables with landscape mode in LaTeX
#cat("\\begin{landscape}\n")
#cat(as.character(formatted_tables[[1]]))
#cat("\n\n")
#cat(as.character(formatted_tables[[2]]))
#cat("\n\n")
#cat(as.character(formatted_tables[[3]]))
#cat("\n\\end{landscape}")

```{r philly-sim-count, echo = FALSE, results='asis'}
#| label: tbl-philly-sim-count
#| tbl-cap: 'Count of Philadelphia area students included in student list purchase that filtered on 2020 high school class and PSAT scores 1070 -- 1180 (Not shown: American Indian/Alaska Native (AI/AN); and Native Hawaiian and other Pacific Islander (NHPI))'

cat(as.character(formatted_tables[[1]]))
```

```{r philly-sim-col-pct, echo = FALSE, results='asis'}
#| label: tbl-philly-sim-col-pct
#| tbl-cap: 'Percent of total prospects purchased from each geomarket, Philadelphia area students included in student list purchase that filtered on 2020 high school class and PSAT scores 1070 -- 1180 (Not shown: AI/AN, NHPI)'

cat(as.character(formatted_tables[[2]]))
```

```{r philly-sim-row-pct, echo = FALSE, results='asis'}
#| label: tbl-philly-sim-row-pct
#| tbl-cap: 'Racial composition of purchased prospects from each Geomarket, Philadelphia area students included in student list purchase that filtered on 2020 high school class and PSAT scores 1070 -- 1180 (Not shown: AI/AN, NHPI)'

cat(as.character(formatted_tables[[3]]))
```

```{r fig-nyc-geomarkets, echo=FALSE, fig.cap="Geomarkets in New York City Area, NY state", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','nyc_geomarkets.png'))
```

```{r fig-asu-order-448922, echo=FALSE, fig.cap="Partial order summary from a College Board student list purchase", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','asu_order_448922_Redacted.pdf'))
```

```{r fig-socal-hispanic, echo=FALSE, fig.cap="Screenshot of Los Angeles area Geomarkets, showing 2020 percent Hispanic by census tract", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','socal_hispanic.png'))
```

```{r fig-socal-med-inc, echo=FALSE, fig.cap="Screenshot of Los Angeles area Geomarkets, showing 2020 median household income by census tract", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','socal_med_inc.png'))
```

```{r fig-philly-black-prospects, echo=FALSE, fig.cap="Screenshot of Philadelphia area Geomarkets, markers are high schools and marker size a function of number of Black student profiles purchased (2020 HS class, PSAT scores 1070 -- 1180)", fig.width=10, fig.height=7, out.extra='angle=90', out.width='110%'}
knitr::include_graphics(path = file.path('.','..','results','figures','philly_black_prospects.png'))
```

>>>>

```{=html}
<!--  
\thispagestyle{empty} 
\newpage 
\setcounter{page}{1}
Please submit a detailed 2-page précis that includes a brief discussion of the topic, pertinent literature and
theory, research methods, findings, and study significance

@@

HOW TO CONNECT WITH THE THEME
- sociology of organizations knows that regulating an industry requires understanding the organizational field, which includes third-party vendors and products. these third party vendors and agents structure college access in ways that benefit historically privileged groups and contribute to accumulation of capital
- we must understand the field before we can regulate it; CFPB has 
Education policy has been dominated by economists. economics doesn't have the theoretical concepts necessary 
-->
```

@@

# ABSTRACT

College Board (CB) "Geomarkets" carve metropolitan areas into smaller geographies. Geomarkets are a "search filter" in the CB "student list" product that enables colleges to acquire prospect contact information. If Geomarket borders are correlated with race and class, using Geomarkets to target prospects may amplify inequality. We address two questions: What is the socioeconomic and racial variation between Geomarkets? How does the socioeconomic and racial composition of included versus excluded prospects vary if student list purchases filter on Geomarkets? We answer RQ1 using Census data and RQ2 using data from actual student list purchases.

@@

**BACKGROUND.** In 1983, Zemsky and Oedel authored *The Structure of College Choice*. Based on an analysis of 1980 SAT score-sending behavior and published by *The College Board* (CB), @RN4982 developed the Market Segment Model and created "Geomarkets." The Market Segment Model predicts how student demand for a particular college varies across local geographic markets based on the characteristics of students in those local markets.

Geomarkets are geographic areas that divide states and large metropolitan areas into "community-based enrollment markets" [@RN4982, p. 14] for recruiting purposes. Figure 1 shows New York City area Geomarkets. Geomarkets were incorporated into the CB "student list" product named Student Search Service. Student lists have been the primary source of lead generation in U.S. higher education for over 50 years [@belkin2019-studata; @list_biz]. Lists contain contact information for prospective students. The Student Search Service database consists of CB test-takers. Colleges control which prospect profiles they purchase by selecting search filters, such as high school graduation year, SAT score, AP score, state, etc. Geomarket filters enable colleges to target prospects living in particular Geomarkets.

**THEORY & RQs.** We analyze the Market Segment Model and Geomarkets using scholarship on quantification [e.g., @RN5003], particularly the discussions of correlation and homophily from @RN4975. Predictive analytics are based on correlation. The 1980 SAT score-sending data analyzed by @RN4982 can be conceived as "training data," which was used to define four "market segments" of students: "local" students, who send most SAT scores to colleges in their Geomarket; "in-state" students, who primarily send scores to colleges in their state; "regional" students, who send most scores to colleges outside their state but in their region (e.g., New England); and "national" students. @RN4982 analyzed the student characteristics correlated with student market segment and concluded that student demand was primarily a function of class. The overarching thesis of the Market Segment Model is homophily, the idea that similar colleges compete with one another for similar students. @RN4982 [p. 42-45] states that selective colleges primarily draw from students in the "regional" and "national" segments and, therefore, their student list purchases should focus on Geomarkets with large numbers of affluent, highly educated households.

When we use data on past correlations to make recommendations about the future, we amplify the effects of historic structural inequality [@RN4794]. A snapshot of student demand in 1980, itself a consequence of historical structural inequality, became inscribed into the Student Search Service product. The result is a supply-side that reinforces structural inequalities observed on the demand-side. We address two research questions:

@@

**SIGNIFICANCE.** Academic scholarship on enrollment management tends to focus on students (micro), policies (macro), and organizations (meso). Scholars have not analyzed third-party classification systems and products that colleges utilize to classify prospective students. CB Geomarkets, utilized in the CB Student Search Service, offers one such case.

**DATA SOURCES**. To answer RQ1, we utilize census tract-level data from the 1980 Decennial Census, 2000 Decennial Census, and 2020 5-year American Community Survey (ACS). Variables of interest include: race/ethnicity; mean and median household income; percent of households below poverty line; and BA attainment of people aged 25+. We obtained shapefiles for CB Geomarkets from a 2012 *R-bloggers* post. We assigned census tracts to Geomarkets.

To answer RQ2, we issued public records requests about student lists purchased by 14 public universities from 2016 through 2020. For 414 CB student list purchases, associated with about 2.6 million prospects, we received both (1) the order summary -- showing which search filters were utilized -- and (2) the de-identified prospect-level data, including race, ethnicity, zip code, and high school. Figure 2 shows the partial order summary for a purchase that filtered on the 2020 high school graduating class, PSAT scores of 1070 to 1180, and selected Geomarkets.

@@

**PRELIMINARY FINDINGS**. For RQ1, Table 1 shows racial and socioeconomic characteristics for Geomarkets around Chicago. Figure 3 is a screenshot of an interactive choropleth map of Los Angeles, showing Geomarket borders and the tract-level %Hispanic in 2020. Figure 4 shows tract-level median household income around Los Angeles.

Results for RQ2 show how many purchased prospects from each racial/ethnic group live in each Geomarket. These prospects would have been excluded had the purchase filtered on Geomarkets and not selected that Geomarket. Table 2, Table 3, and Table 4, respectively, show counts, column percentages, and row percentages for the Philadelphia metropolitan area, based on a purchase that filtered on students from the 2020 high school class who scored between 1070 and 1180 on the PSAT. Table 3 shows that "PA5--Philadelphia County" contains 15% of all purchased prospect profiles but 44% of Black prospects, indicating that opportunity for Black students is sensitive to the inclusion of this Geomarket. Figure 5 is a screenshot of a (very) preliminary interactive map for this purchase, with marker size indicating the number of purchased prospect profiles who attended that high school and separate markers by race/ethnicity.

**PRESENTATION PLAN**. I will provide a succinct background about CB Geomarkets and the Student Search Service product, but I will omit discussion of theory. I will show actual student list purchases that target the Chicago-land area but exclude the City of Chicago Geomarket. Next, I will show which prospects are excluded by this decision. Finally, I will engage participants in a discussion about contemporary third-party products that classify students on behalf of colleges. How can practitioners and policymakers effectively interrogate these products to avoid unintentional exclusion along racial and class dimensions?

@@
